A Theory for Distributed Database Transactions

### 2019-06-14 2018_Transactions/articleTransactions.tex
language and explanation

\begin{abstract}
	A shared data space can be problematic in globally distributed information systems,
	because it can become a performance bottleneck or a single point of failure.
	Yet, transaction theory as we know it assumes a shared data space.
	Several solutions exist, such as nested transactions and distributed transactions.
	Practitioners tend to avoid them when possible becaus the are complicated and error prone.

@@

\section{Introduction}
\label{sct:Introduction}
	A distributed information system (DIS) serves its users with things such as
	registering data, guarding consistency, protecting that data,
	and providing it accurately and meaningfully for its intended purpose to various users.
	These users want predictable behaviour, even when they change a shared state,
	e.g. when reserving a seat in a global flight reservation system like Galileo.
	Although we cannot always avoid sharing a state,
	we usually can avoid sharing an entire database.
	Globally distributed applications such as Netflix and LinkedIn cannot afford to rely on a shared database for obvious reasons of performance and reliability.
	Many cloud applications use distributed storage models instead.
	Examples are event stores (e.g.\ Kafka), triple stores (e.g.\ semantic web), graph databases (e.g. Neo4J) and others.
	Many so-called NoSQL databases lack functionality for transactions,
	leaving the problems for the user to solve.
	As a result, architectural patterns exist for the purpose of doing transactions in a distributed setting.
	Some of these patterns simply circumvent the problem.
	Others will isolate transactionality into a smaller sub-problem.
	Still others introduce locking mechanisms to do the job.

	Our work is inspired by the observation that the merge of Git corresponds to the commit that we know from transactional databases.
	We show this correspondence by expressing the characteristic properties of transactions in a theory of time.
	For this purpose we build on a well-known theory of time~\cite{DBLP:journals/ipl/FosterCWZ18},
	which is generally applicable,
	well-established, and founded in a solid mathematical structure~\cite{Foster14}.
	This means we have to redefine Atomicity, Consistency, Isolation, and Durability in terms of events.
	Informally, {\em Atomicity} means that a sequence of primitive actions can be executed as though it were a single action.
	{\em Consistency} means that integrity rules are warranted.
	{\em Isolation} means that actions can be done without any knowledge of what other users are doing.
	{\em Durability} means that history does not change.
	The result of this work contributes to the body of foundational theory of information systems
	by defining transactions in a theory of time.
	By showing that transactions come quite naturally from an event perspective,
	we give formal substance under the observation that event systems like Git do not need new features to behave transactionally.

@@

	We use variables $e$, $e_1$, and $e_2$ to represent events.
	Let $\mathbb E$ represent the set of all conceivable events.
%	We also introduce the notion of {\em history structure}
%	to represent all events that have happened up to some point in time and the precedences between those events.
%	In this paper, variables $h$, $h'$, $h_1$, and $h_2$ represent history structures.

@@

\section{Traces}
\label{sct:Traces}
	Transactions leave traces in the history of a DIS, so let us define traces.
	For this purpose we have chosen definitions that satisfy trace algebra
	and represent traces in discrete-event systems.
	Other interpretations of the same trace algebra comprise real-time systems and hybrid systems,
	which is proven by Foster~\cite{Foster17b}.
	We prove (TBD) that our definitions comply with Foster's axioms of trace algebra,
	but stick to one interpretation (discrete event systems).

@@

	Let us define prefix for traces ($\leq$), to correspond closely to the prefix for histories
	(as given in definition~\ref{def:lies in the past}):
\begin{definition}[Prefix]
\label{def:Prefix of traces}
\item $\begin{array}[t]{rcl}
	(e,e_1)_\id{prec}\leq(e,e_2)_{\id{prec}'}&\Leftrightarrow&\id{prec}\leq\id{prec}'\ \wedge\ e_1\ \id{prec}'\ e_2
\end{array}$
\end{definition}
	We have defined this operator such that it satisfies Foster's algebraic definition of prefix~\cite{Foster17b}.
	(proof needed)
\begin{eqnarray}
	x\leq y&\Rightarrow&\exists z:\ x+z=y\label{def:Prefix}
\end{eqnarray}
	The prefix allows us to define subtraction of traces, to have a way to ``cut off'' an ``older'' part of a trace.
\begin{definition}[Subtract]
\label{def:Subtract}
\item $\begin{array}[t]{rcll}
	(e,e_1)_\id{prec}-(e,e_2)_{\id{prec}'}&=&(e_1,e_2)_{\id{prec}'}&\text{if }(e,e_1)_\id{prec}\leq(e,e_2)_{\id{prec}'}\\
	   &=&0&\text{otherwise}
\end{array}$
\end{definition}
	This definition satisfies Foster's definition of subtract (eqn.~\ref{axi:Foster subtract})
\begin{equation}
\label{axi:Foster subtract}
\begin{array}[t]{rcll}
	y-x&=&z&\text{if }x\leq y\ \wedge\ \exists z: y=x+z\\
	   &=&0&\text{otherwise}
\end{array}
\end{equation}

\section{State}
\label{sct:State}
	Typically, distributed information systems are stateful.
	Every event in the history of a DIS may cause a different (local) state.
	(The distributed nature of a DIS means that states are local.)
	Let $\mathbb S$ be the set of all conceivable states.
	So we assume a function $\id{state}_\id{prec}:\mathbb E\rightarrow\mathbb S$ that assigns one state to every event $e$ in history $\id{prec}$.
	State $\id{state}_\id{prec}(e)$ represents the state immediately after $e$ has happened.
	We assume an empty state, $\emptyset$, which is the state immediately before an initial event has happened.
	The global state of the system, if relevant, can be computed by $\{\id{state}_\id{prec}(e) | e\in\id{current}(\id{prec})\}$.
	Locally we work with one specific current event, $e$, with $\id{state}_\id{prec}(e)$ representing the local current state.

@@

	An action is denoted as $\id{act}_{e_1,e_2,\Delta}$, which is a mapping from one DIS to another
\begin{equation}
\begin{array}{rl}
	\multicolumn{2}{l}{e_2\not\in\id{events}(\id{prec})\ \wedge\ (e_1,e_2)\in\id{prec}'^*\ \Rightarrow}\\
	&\id{act}_{e_1,e_2,\Delta}(\id{prec},\delta_\id{prec})\ =\ (\id{prec}',\delta'_{\id{prec}'})\\
	&\text{where}\\
	&\ \ \begin{array}[t]{rcll}
		\delta'_{\id{prec}'}(e,e')&=&\Delta&\text{if }(e,e')=(e_1,e_2)\\
		\delta'_{\id{prec}'}(e,e')&=&\delta_\id{prec}(e,e')&\text{otherwise}\\
		\end{array}
\end{array}
\end{equation}
	The state function can be computed recursively as follows:
\begin{equation}
\begin{array}{rcll}
	\id{state}_\id{prec}(e)&=&\id{state}_\id{prec}(e_1)\oplus\delta(e_1,e)&\text{if }(e_1,e)\in\id{prec}\\
	                       &=&\emptyset&\text{otherwise}
\end{array}
\end{equation}

	Durability is meant to express that an action preserves history.
\begin{definition}[Durability]
\item An action $\id{act}:\mathbb H\rightarrow\mathbb H$ is durable iff for every $h\in\mathbb H$:
\label{def:Durability}
\begin{eqnarray}
	\id{prec}^*\subseteq\id{prec}'^*
\end{eqnarray}
\end{definition}

	An action $\id{act}_\id{prec}(e_1,e_2)$ is performed in isolation if it does not affect any other action in the system.
	This means that the only event in trace $(e_1,e_2)_\id{prec}$ that precedes events outside that trace is $e_1$.
	Likewise, the only event in trace $(e_1,e_2)_\id{prec}$ that is preceded by events outside that trace is $e_2$.
	To define this, we need an auxiliary function
	$\id{atomize}:\mathbb E\times\mathbb E\times\mathbb H\rightarrow\mathbb H$,
	which replaces a trace $(e_1,e_2)_\id{prec}$ by the pair $(e_1,e_2)$
	in a history $\id{prec}$.
\begin{eqnarray}
	\begin{array}[t]{r@{}l@{}l}
	\multicolumn{3}{l}{\id{atomize}(e_1,e_2,\id{prec})}\\
	   \ =\ \id{prec}-&(&\{ (e_1,e)\in\id{prec}| e_1\ \id{prec}^*\ e\ \wedge\ e\ \id{prec}^*\ e_2\}\ \cup\\
			  & &\{ (e,e_2)\in\id{prec}| e_1\ \id{prec}^*\ e\ \wedge\ e\ \id{prec}^*\ e_2\}\\
			  &)\\
		\cup&&\{ (e_1,e_2) \}\\
	\end{array}
\end{eqnarray}
	We can now define isolation as the property that the traces of the "atomized" history
	are the same as the atomization of all traces in that history.
\begin{definition}[Isolation]
\item An action $\id{act}_\id{prec}(e_1,e_2)$ works in isolation iff:
\label{def:Isolation}
\begin{eqnarray}
	(\id{atomize}(e_1,e_2,\id{prec}))^*&=&\id{atomize}(e_1,e_2,\id{prec}^*)
\end{eqnarray}
\end{definition}

	Atomicity expresses that a sequence of multiple actions can be treated as one.

\begin{definition}[Atomic]
\item An action $\id{act}_\id{prec}(e_1,e_2)$ is atomic iff
\label{def:Atomicity}
\end{definition}

>>>>

\begin{abstract}
	A shared data space can be problematic in globally distributed information systems,
	because it can become a performance bottleneck or a single point of failure.
	Yet, transaction theory as we know it assumes a shared data space.
	Several solutions exist, such as nested transactions and distributed transactions.
	Practitioners tend to avoid them when possible because they are complicated and error-prone.

@@

\section{Introduction}
\label{sct:Introduction}
	A distributed information system (DIS) serves its users with things such as
	registering data, guarding consistency, protecting that data,
	and providing it accurately and meaningfully for its intended purpose to various users.
	These users want predictable behaviour, even when they change a shared state,
	e.g. when reserving a seat in a global flight reservation system like Galileo or Amadeus.
	Although we cannot always avoid sharing a state,
	we usually can avoid sharing an entire database.
	Globally distributed applications such as Netflix and LinkedIn cannot afford to rely on a shared database for obvious reasons of performance and reliability.
	Many cloud applications use distributed storage models instead.
	Examples are event stores (e.g.\ Kafka), triple stores (e.g.\ semantic web), graph databases (e.g. Neo4J) and others.
	Many so-called NoSQL databases lack functionality for transactions,
	leaving the problems for the user to solve.
	As a result, architectural patterns exist for the purpose of doing transactions in a distributed setting.
	Some of these patterns simply circumvent the problem.
	Others will isolate transactionality into a smaller sub-problem.
	Still others introduce locking mechanisms to do the job.

	Our work is inspired by the observation that the merge of Git corresponds to the commit that we know from transactional databases.
	We show this correspondence by expressing the characteristic properties of transactions in a theory of time.
	For this purpose we build on a well-known theory of time~\cite{DBLP:journals/ipl/FosterCWZ18},
	which is generally applicable,
	well-established, and founded in a solid mathematical structure~\cite{Foster14}.
	This means we have to redefine Atomicity, Consistency, Isolation, and Durability in terms of events.
	Informally, {\em Atomicity} means that a sequence of primitive actions can be executed as though it were a single action.
	{\em Consistency} means that integrity rules are warranted.
	{\em Isolation} means that actions can be done without any knowledge of what other users are doing.
	{\em Durability} means that history does not change.

	The result of this work contributes to the body of foundational theory of information systems
	by defining transactions in a theory of time.
	By showing that transactions come quite naturally from an event perspective,
	we give formal substance under the observation that event systems like Git do not need new features to behave transactionally.

@@

	We use variables $e$, $e'$, $e_1$, and $e_2$ to represent events.
	Let $\mathbb E$ represent the set of all conceivable events.

@@

\section{Traces}
\label{sct:Traces}
	Transactions leave {\em traces} in the history of a DIS, so let us define these traces.
	For this purpose we have chosen definitions that satisfy trace algebra
	and represent traces in discrete-event systems.
	Other interpretations of the same trace algebra comprise real-time systems and hybrid systems,
	which has been  proven by Foster~\cite{Foster17b}.
	We prove (TBD) that our definitions comply with Foster's axioms of trace algebra,
	but stick to one interpretation (discrete event systems).

@@

	Let us define prefix for traces ($\leq$), to correspond closely to the prefix for histories
	(as given in definition~\ref{def:lies in the past}):
\begin{definition}[Prefix]
\label{def:Prefix of traces}
\item $\begin{array}[t]{rcl}
	(e,e_1)_\id{prec}\leq(e,e_2)_{\id{prec}'}&\Leftrightarrow&\id{prec}\leq\id{prec}'\ \wedge\ e_1\ \id{prec}'\ e_2
\end{array}$
\end{definition}
	We have defined this operator such that it satisfies Foster's algebraic definition of prefix~\cite{Foster17b}.
	(proof needed)
\begin{eqnarray}
	x\leq y&\Rightarrow&\exists z:\ x+z=y\label{def:Prefix}
\end{eqnarray}
	The prefix allows us to define subtraction of traces, to have a way to ``cut off'' an ``older'' part of a trace.
\begin{definition}[Subtract]
\label{def:Subtract}
\item $\begin{array}[t]{rcll}
	(e,e_1)_\id{prec}-(e,e_2)_{\id{prec}'}&=&(e_1,e_2)_{\id{prec}'}&\text{if }(e,e_1)_\id{prec}\leq(e,e_2)_{\id{prec}'}\\
	   &=&0&\text{otherwise}
\end{array}$
\end{definition}
	Definition~\ref{def:Subtract} satisfies Foster's definition of subtract:
\begin{equation}
\label{axi:Foster subtract}
\begin{array}[t]{rcll}
	y-x&=&z&\text{if }x\leq y\ \wedge\ \exists z: y=x+z\\
	   &=&0&\text{otherwise}
\end{array}
\end{equation}

\section{State}
\label{sct:State}
	Typically, distributed information systems are stateful.
	Every event in the history of a DIS may cause a different (local) state.
	(The distributed nature of a DIS means that states are local.)
	Let $\mathbb S$ be the set of all conceivable states.
	So we need a function $\id{state}_\id{prec}:\mathbb E\rightarrow\mathbb S$ that assigns one state to every event $e$ in history $\id{prec}$.
	State $\id{state}_\id{prec}(e)$ represents the state immediately after $e$ has happened.
	If the history $\id{prec}$ is obvious from the context, we may write $\id{state}(e)$ rather than $\id{state}_\id{prec}(e)$.
	We assume an empty state, $\emptyset$, which is the state immediately before an initial event has happened.
	The global state of the system, if relevant, can be computed by $\{\id{state}(e) | e\in\id{current}(\id{prec})\}$.
	Locally we work with one specific current event, $e$, with $\id{state}_\id{prec}(e)$ representing the local current state.

@@

	An action is denoted as $\id{act}_{e_1,e_2,\Delta}$, which is a mapping from one DIS to another
\begin{equation}
\label{def:action}
\begin{array}{rl}
	\multicolumn{2}{l}{e_2\not\in\id{events}(\id{prec})\ \wedge\ (e_1,e_2)\in\id{prec}'^*\ \Rightarrow}\\
	&\id{act}_{e_1,e_2,\Delta}(\id{prec},\delta_\id{prec})\ =\ (\id{prec}',\delta'_{\id{prec}'})\\
	&\text{where}\\
	&\ \ \begin{array}[t]{rcll}
		\delta'_{\id{prec}'}(e,e')&=&\Delta&\text{if }(e,e')=(e_1,e_2)\\
		\delta'_{\id{prec}'}(e,e')&=&\delta_\id{prec}(e,e')&\text{otherwise}\\
		\end{array}
\end{array}
\end{equation}
	The state function can be computed recursively as follows:
\begin{equation}
\begin{array}{rcll}
	\id{state}_\id{prec}(e)&=&\id{state}_\id{prec}(e_1)\oplus\delta(e_1,e)&\text{if }(e_1,e)\in\id{prec}\\
	                       &=&\emptyset&\text{otherwise}
\end{array}
\end{equation}

	Now let us define the classical characteristic properties of transactions.
	We start with the easiest one: durability.
	This property expresses that an action must preserve history.
\begin{definition}[Durability]
\item An action $\id{act}_{e_1,e_2,\Delta}$ is durable iff:
\label{def:Durability}
\[\begin{array}{rl}
	&\id{act}_{e_1,e_2,\Delta}(\id{prec},\delta_\id{prec})\ =\ (\id{prec}',\delta'_{\id{prec}'})\\
	\Rightarrow\\
	&\id{prec}^*\subseteq\id{prec}'^*\ \wedge\ \delta_\id{prec}\subseteq\delta'_{\id{prec}'}
\end{array}\]
\end{definition}
	Note that durability is already satisfied in the definition of action (equation~\ref{def:action}). (proof needed)
	Such is not the case with isolation.
	An action $\id{act}_\id{prec}(e_1,e_2)$ is isolated, i.e. performed in isolation, if it does not affect any other action in the system.
	This means that the only event in trace $(e_1,e_2)_\id{prec}$ that precedes events outside that trace is $e_1$.
	Likewise, the only event in trace $(e_1,e_2)_\id{prec}$ that is preceded by events outside that trace is $e_2$.
	To define this, we need an auxiliary function
	$\id{atomize}:\mathbb E\times\mathbb E\times\mathbb H\rightarrow\mathbb H$,
	which replaces a trace $(e_1,e_2)_\id{prec}$ by the pair $(e_1,e_2)$
	in a history $\id{prec}$.
\begin{eqnarray}
	\begin{array}[t]{r@{}l@{}l}
	\multicolumn{3}{l}{\id{atomize}(e_1,e_2,\id{prec})}\\
	   \ =\ \id{prec}-&(&\{ (e_1,e)\in\id{prec}| e_1\ \id{prec}^*\ e\ \wedge\ e\ \id{prec}^*\ e_2\}\ \cup\\
			  & &\{ (e,e_2)\in\id{prec}| e_1\ \id{prec}^*\ e\ \wedge\ e\ \id{prec}^*\ e_2\}\\
			  &)\\
		\cup&&\{ (e_1,e_2) \}\\
	\end{array}
\end{eqnarray}
	We can now define isolation as the property that the traces of the ``atomized'' history
	are the same as the atomization of all traces in that history.
\begin{definition}[Isolation]
\item An action $\id{act}_{e_1,e_2,\Delta}$ works isolated in history $\id{prec}$ iff:
\label{def:Isolation}
\begin{eqnarray}
	(\id{atomize}(e_1,e_2,\id{prec}))^*&=&\id{atomize}(e_1,e_2,\id{prec}^*)
\end{eqnarray}
\end{definition}

	Atomicity is related to isolation because it expresses that a sequence of multiple actions can be treated as one.

\begin{definition}[Atomic]
\item An action $\id{act}_{e_1,e_2,\Delta}$ is atomic in history $\id{prec}$ iff:
\[\begin{array}{rl}
	&\id{act}_{e_1,e_2,\Delta}(\id{prec},\delta_\id{prec})\ =\ (\id{prec}',\delta'_{\id{prec}'})\\
	\Rightarrow\\
	&\id{atomize}(e_1,e_2,\id{prec})\ \leq\ \id{prec}'
\end{array}\]
\label{def:Atomicity}
\end{definition}
